---
title: "Data_transformation"
author: "Goutham"
date: "10/18/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(nycflights13)
```


## Little practice before the exercises...
```{r}
?flights

flights
View(flights)  # To view the entire dataset
dim(flights)

filter(flights, day == 1, month == 1) %>% View()

jan1 <- filter(flights, day == 1, month == 1)
jan1

(nov_dec <- filter(flights, month %in% c(11,12)))
(nov_dec <- filter(flights, month == 11 | month == 12)) # Another way of writing above

#Demorgans law:
# !(x | y) = !x & !y
#! (x & y ) = !x | !y



```


### 5.3.4 Filtering rows with filter()

1. Find all flights that

1a. Had an arrival delay of two or more hours
```{r}
filter(flights, arr_delay >= 120)
```

1b. Flew to Houston (IAH or HOU)
```{r}
filter(flights, dest %in% c("IAH", "HOU"))
```

1c. Were operated by United, American, or Delta
```{r}
filter(flights, carrier %in% c("UA", "AA","DL"))
```

1d. Departed in summer (July, August, and September)
```{r}
filter(flights, month %in% c(7:9))
```

1e. Arrived more than two hours late, but didn’t leave late
```{r}
filter(flights, arr_delay > 120 & dep_delay <= 0)
```

1f. Were delayed by at least an hour, but made up over 30 minutes in flight
```{r}
filter(flights, dep_delay >= 60 & dep_delay - arr_delay > 30)
```

1g. Departed between midnight and 6am (inclusive)
```{r}
filter(flights, dep_time <= 600 | dep_time >= 2400) %>% dim()

summary(flights$dep_time)
```


2. Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?
```{r}
?between

between(1:12, 7, 9)
```


3. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?
```{r}
filter(flights, is.na(dep_time))
```

4. Why is NA ^ 0 not missing? Why is NA | TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)
```{r}
NA ^ 0
NA | TRUE
FALSE & NA
NA * 0
```
NA ^ 0 is 1 because any x ^ 0 = 1
Na | TRUE is TRUE. Missing vale is T or F and T or F | T is always T
FALSE & NA is NA because anything false with & operator is always a false.
NA * 0 is NA because....?


### 5.3.1 Arrange rows with arrange()

1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).
```{r}
arrange(flights, desc(is.na(dep_time)))


(df <- tibble(x = c(5, 2, NA), y = c(1, NA, 4)))
arrange(df, desc(is.na(y)))

?arrange
```

2. Sort flights to find the most delayed flights. Find the flights that left earliest.
```{r}
# summary(flights$dep_delay)

arrange(flights, desc(dep_delay))
arrange(flights, dep_delay)
```

3. Sort flights to find the fastest flights.
```{r}
arrange(flights, distance %% air_time)
```

4. Which flights travelled the longest? Which travelled the shortest?
```{r}
arrange(flights, air_time)
arrange(flights, desc(air_time))
```


### 5.4.1 Select columns with select()

1. Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.
```{r}
names(flights)
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, 4, 6, 7, 9)
select(flights, dep_time:arr_delay, - sched_dep_time, - sched_arr_time)
select(flights, starts_with("dep"), starts_with("arr"))
select(flights, contains("delay"), contains("time"), - carrier, -sched_dep_time, -sched_arr_time, -air_time, - time_hour)
```

2. What happens if you include the name of a variable multiple times in a select() call?
```{r}
select(flights, arr_time, arr_time, arr_time, dep_time, dep_time)
```
R only optputs the variable once. The duplicates are ignored.

3. What does the one_of() function do? Why might it be helpful in conjunction with this vector?
```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, one_of(vars))

?select
```
one_of function amtches the names in the character vector to the variables in the dataframe.

4. Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?
```{r}
select(flights, contains("TIME"))
select(flights, contains("TIME", ignore.case = FALSE))
```
See help for helper function. In the arguments of the helper function, the default is to ignore the case of the string. If the ignore.case is set to FALSE, the result looks more like what we would expect.






















